### 2021-10-27 text/outline/outline.tex
add fig 1

Our work makes two main contributions. First, we conduct an in-depth study of the thresholding method from empirical and theoretical perspectives. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian model. Second, we introduce a new method for estimation and inference in single-cell CRISPR screens that accounts for the measurement error inherent in the experiment. The method, called \textit{GLM-EIV} (generalized linear model with errors-in-variables), implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic. Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.9\linewidth]{analysis_challenges}
	\caption{\textbf{Overview of experimental design and analysis challenges}: \textbf{a,} Experimental design. For a given perturbation (e.g., the perturbation represented in yellow), we partition the cells into two groups: those that received the perturbation, and those that did not receive the perturbation. For a given gene, we conduct a differential expression analysis across the two groups of cells, yielding an estimate of the impact of the given perturbation on the given gene. \textbf{b,} DAG representing the variables in the analysis. The perturbation (unobserved) affects both gene expression and gRNA expression; technical factors (e.g., batch, sequencing depth, etc.) act as nuisance variables. The target of inference is the effect of the perturbation on gene expression (denoted with question mark). \textbf{c,} Schematic illustrating ``background reads.'' The gRNA modality has a nonzero, ``background read'' distribution even in the absence of a perturbation, complicating the assignment of perturbations to cells. \textbf{d}, Example data for a given perturbation-gene pair. Notice that (i) the perturbations are unobserved, and (ii) the gene and gRNA expression data take the form of discrete counts.}
	\label{analysis_challenges}
\end{figure}

@@

In this section we study thresholded regression from empirical and theoretical perspectives, highlighting several fundamental challenges of the method. Let $n \in \N$ be the number of cells assayed in the experiment (typically, $n \approx 100,000 - 250,000$). Consider a single perturbation and a single gene. For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced in the cell; let $g_i \in \N$ be the number of gRNA transcripts sequenced in the cell; let $l^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (the library size) in the cell; and finally, let $z_i \in \R^{d-1}$ be the cell-specific vector of technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.). The thresholding method is defined as follows:
\begin{itemize}
\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i \geq c, \\ \hat{p}_i = 1 \textrm{ if } g_i < c. \end{cases}$$
\item[2.] Assume that $m_i$ is related to $\hat{p}_i, l_i,$ and $z_i$ through the following GLM:
\begin{equation*}
\begin{cases}
m_i | (\hat{p}_i, z_i, l_i) \sim \textrm{NB}_\theta(\mu_i), \\
\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left( l_i^m\right),
\end{cases}
\end{equation*}
where (i) $NB_\theta(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(l_i^m)$ is an offset term. Fit a GLM to obtain estimates of the parameters.

@@

The biological interpretation for the target of inference $\beta^m_1$ as follows:  $\beta^m_1$ is the log-transformed fold change in gene expression in response to the perturbation, removing the effect of the technical factors. Fold change (obtained by exponentiating $\beta^m_1$) is the ratio of the mean expression in cells that were perturbed to the mean expression in cells that were not perturbed. $\exp(\beta^m_1) = 1$ indicates no change in mean expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively.

@@

We studied the behavior of the thresholding method on real data, uncovering an attenuation bias effect and a bias-variance tradeoff as a function of the selected threshold. We applied the thresholding method to analyze the set of 381 positive control perturbation-gene pairs in the Gasperini dataset. The positive control pairs consisted of perturbations that targeted gene transcription start sites (TSSs) for inhibition. Repressing the TSS of a given gene decreases its expression; therefore, the positive control pairs \textit{a priori} are expected to exhibit a strong, negative log fold change in expression.

To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method using three different choices for the threshold: 1, 5, and 20. We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b). Estimates for log fold change produced by setting the threshold to 1 were smaller in magnitude than those produced by setting the threshold to 5. (Equivalently, estimates for \textit{raw} fold change were closer to the baseline of $1$ for threshold = 1; Figure \ref{thresholding_empirical}a.) Fold change estimates generated by setting the threshold to 5 and setting the threshold to 20 were more concordant; however, the threshold of 20 yielded slightly stronger effect sizes (Figure \ref{thresholding_empirical}b). These results are consistent with an \textit{attenuation bias} phenomenon: small thresholds tend to underestimate the true parameter value, and this effect diminishes gradually as the threshold increases.

We hypothesized that selecting a large threshold comes with the cost of increasing the variance of the estimator.

% \blue{To investigate the impact of threshold selection on the thresholding method, we applied the thresholding method to the set of positive control (i.e., gene-targeting) perturbation-gene pairs in the Gasperini data using three different choices for the threshold: 1, 5, and 20.}

% \blue{Description of panels a-d; threshold = 1 leads to considerable attenuation bias (a); threshold = 5 and threshold = 20 produce similar estimates, though the effect sizes are slightly greater for threshold = 20, suggesting threshold = 5 yields mild attenuation bias (b); the threshold = 20 estimates are more variable than than those of threshold = 5, as evidenced by the smaller $p$-values and wider confidence intervals.}

% \blue{Description of panels e-f: these are the empirical gRNA count distributions for randomly selected gRNAs from the Gasperini and Xie datasets. There does not appear to be a clear location at which to draw the threshold; aside from the initial spike at one (zero not shown), the histograms gradually decrease.}

% \blue{Take-home message: (i) the threshold is a tuning parameter that substantially affects the result; (ii) as the threshold increases, bias seems to decrease and variance seems to increase; (iii) it is not clear where to draw the threshold from gRNA counts alone.}

\begin{figure}[h!]
	\centering
	\includegraphics[width=1\linewidth]{../../figures/thresholding_empirical/plot}
	\caption{\textbf{Empirical challenges of thresholded regression.}}
	\label{thresholding_empirical}
\end{figure}
\newpage

>>>>

Our work makes two main contributions. First, we conduct an investigation of the thresholding method from empirical and theoretical perspectives. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian model. Second, we introduce a new method for estimation and inference in single-cell CRISPR screens that accounts for the measurement error inherent in the experiment. The method, called \textit{GLM-EIV} (generalized linear model with errors-in-variables), implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic. Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=1\linewidth]{../../figures/analysis_challenges/plot.pdf}
	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design. For a given perturbation (e.g., the perturbation represented in yellow), we partition the cells into two groups: those that received the perturbation, and those that did not receive the perturbation. For a given gene, we conduct a differential expression analysis across the two groups of cells, yielding an estimate of the impact of the given perturbation on the given gene. \textbf{b,} DAG representing the variables in the analysis. The perturbation (unobserved) affects both gene expression and gRNA expression; technical factors (e.g., batch, sequencing depth, etc.) act as nuisance variables. The target of inference is the effect of the perturbation on gene expression (denoted with question mark). \textbf{c,} Schematic illustrating ``background reads.'' The gRNA modality has a nonzero, ``background read'' distribution even in the absence of a perturbation, complicating the assignment of perturbations to cells. \textbf{d}, Example data for a given perturbation-gene pair. Notice that (i) the perturbations are unobserved, and (ii) the gene and gRNA expression data take the form of discrete counts.}
	\label{analysis_challenges}
\end{figure}

@@

In this section we study thresholded regression from empirical and theoretical perspectives, highlighting several limitations of the method. Let $n \in \N$ be the number of cells assayed in the experiment (typically, $n \approx 100,000 - 250,000$). Consider a single perturbation and a single gene. For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced in the cell; let $g_i \in \N$ be the number of gRNA transcripts sequenced in the cell; let $l^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (the library size) in the cell; and finally, let $z_i \in \R^{d-1}$ be the cell-specific vector of technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.). The thresholding method is defined as follows:
\begin{itemize}
\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i \geq c, \\ \hat{p}_i = 1 \textrm{ if } g_i < c. \end{cases}$$
\item[2.] Assume that $m_i$ is related to $\hat{p}_i, l_i,$ and $z_i$ through the following GLM:
\begin{equation*}
\begin{cases}
m_i | (\hat{p}_i, z_i, l_i) \sim \textrm{NB}_\theta(\mu_i), \\
\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left( l_i^m\right),
\end{cases}
\end{equation*}
where (i) $NB_\theta(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(l_i^m)$ is an offset term. Fit a GLM to obtain estimates of the parameters.

@@

The biological interpretation for the target of inference $\beta^m_1$ as follows:  $\beta^m_1$ is the log-transformed fold change in gene expression in response to the perturbation, removing the effect of the technical factors. Fold change (obtained by exponentiating $\beta^m_1$) is the ratio of the mean expression in cells that were perturbed to the mean expression in cells that were not perturbed. $\exp(\beta^m_1) = 1$ indicates no change in mean expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively (accounting for technical factors).

@@

We examined the behavior of the thresholding method on real data and uncovered attenuation bias and bias-variance tradeoff effects. We applied the thresholding method to analyze the set of 381 positive control perturbation-gene pairs in the Gasperini dataset. The positive control pairs consisted of perturbations that targeted gene transcription start sites (TSSs) for inhibition. Repressing the TSS of a given gene decreases its expression; therefore, the positive control pairs \textit{a priori} are expected to exhibit a strong, negative log fold change in expression.

To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method on the positive control data using three different thresholds: 1, 5, and 20. We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b). Estimates for log fold change produced by threshold = 1 were smaller in magnitude than those produced by threshold = 5. (Equivalently, estimates for \textit{raw} fold change were closer to the baseline of $1$ for threshold = 1; Figure \ref{thresholding_empirical}a.) Estimates produced by threshold = 5 and threshold = 20 were more concordant, but threshold = 20 yielded slightly larger effect sizes (Figure \ref{thresholding_empirical}b).

We reasoned that thresholded regression underestimated effect sizes on the positive control pairs, especially for threshold = 1 (an example of \textit{attenuation bias}). For a given perturbation, the vast majority ($>99\%$) of cells are unperturbed. This imbalance leads to an asymmetry: misclassifying \textit{unperturbed} cells as \textit{perturbed} is intuitively ``worse'' than misclassifying \textit{perturbed} cells as \textit{unperturbed}. Misclassified \textit{unperturbed} cells contaminate the set of truly perturbed cells, leading to attenuation bias; by contrast, misclassified \textit{perturbed} cells are swamped in number and ``neutralized'' by the truly \textit{unperturbed} cells. Setting the threshold to a large number reduces the unperturbed-to-perturbed misclassification rate, decreasing bias.

We hypothesized, however, that the reduction in bias conferred by selecting a large threshold incurs the cost of increasing the variance of the estimator. To investigate, we compared $p$-values and confidence intervals produced by threshold = 5 and threshold = 20 for the target of inference $\beta^m_1$. We found that threshold = 5 yielded smaller (i.e., more significant) $p$-values and narrower confidence intervals than did threshold = 20 (Figure \ref{thresholding_empirical}c-d). We concluded that thresholded regression has a bias-variance tradeoff: as the threshold increases, bias decreases and variance increases.

Finally, to determine whether there is an ``obvious'' location at which to draw the threshold, we examined the empirical gRNA count distributions and checked for bimodality. Figures \ref{thresholding_empirical}e and \ref{thresholding_empirical}f display the empirical distribution of a randomly-selected gRNA from the Gasperini and Xie datasets, respectively (counts of $0$ omitted). The distributions peak at $1$ and then taper off gradually; there does not exist a sharp boundary that cleanly separates the perturbed from the unperturbed cells. Overall, we concluded that the thresholding method faces several challenges: (i) the threshold is a tuning parameter that significantly impacts the results; (ii) the threshold mediates an intrinsic bias-variance tradeoff; and (iii) the gRNA count distributions do not imply a clear threshold selection strategy.

\begin{figure}[h!]
	\centering
	\includegraphics[width=1\linewidth]{../../figures/thresholding_empirical/plot}
	\caption{\textbf{Empirical challenges of thresholded regression.} \textbf{a-b,} Fold change estimates produced by threshold = 1 versus threshold = 5 (a) and threshold = 20 versus threshold = 5 (b). The selected threshold substantially impacts the results. \textbf{c-d,} $p$-values (c) and CI widths (d) produced by threshold = 20 versus threshold = 5. The latter threshold yields more confident estimates. \textbf{e-f}, Empirical distribution of randomly-selected gRNA from Gasperini (e) and Xie (f) data. The gRNA data do not appear to imply an obvious threshold selection strategy.}
	\label{thresholding_empirical}
\end{figure}
\newpage
